---
title: "Using StatsBombR to obtain soccer data"
subtitle: "Event level soccer data"
author: "Conor Buckley"
date: 2022-03-18
categories: ["R", "Data"]
tags: ["R", "plot", "ggplot2", "soccer"]
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, class.source = "chunk-style")
```

This post shows how I used the 'StatsBombR' R package to get freely available soccer data provided by StatsBomb. [StatsBomb](https://statsbomb.com/) is a sports data company and they provide free online samples of some of their professional level data. One way to access this free data is through their StatsBombR R package. Using this package I demonstrate how to extract the event level dataset for the 2019 Champions League final between Liverpool and Tottenham Hotspur. 

Firstly, I load the required libraries for this post.

```{r, warning=FALSE}
# Load required libraries
library(StatsBombR)
library(tidyverse)
```

Using the FreeCompetitions() function from StatsBombR I obtain all the competitions that have free data available. For those competitions, I obtain the matches with free data. This results in 40 competitions with a total of 1,096 matches.

```{r, warning=FALSE, results='hide'}
# Get the data via StatsBombR package
comps <- FreeCompetitions() # get the available competitions (40x12)
```
```{r, warning=FALSE}
glimpse(comps)
```

Using the FreeMatches() function from StatsBombR I obtain all the matches that have free data available. This dataset corresponds to this dataset in my [Github repo](https://github.com/buckleco/Portfolio_Data/blob/main/matches.csv), which I use in some other blog posts.

```{r, warning=FALSE, results='hide'}
matches <- FreeMatches(comps) # get the available matches in those competitions (1,096x42)
```
```{r, warning=FALSE}
glimpse(matches)
```

The last step is to extract the event level data for a specific match. For this, I chose the 2019 Champions League final between Liverpool and Tottenham Hotspur. 

```{r, warning=FALSE, results='hide'}
# Match number you want to select from the matches dataframe
selected_match <- 1 # Liv v Spurs is the first entry

# Select Liverpool v Spurs 2019 Champions League Final
match_1 <- get.matchFree(matches[selected_match, ]) # (3,165x120)

# Apply the StatsBomb allclean function to get the base dataset
# The allclean() function cleans and enriches the dataset for use
match_1_clean <- allclean(match_1) # (3,165x157)
```
```{r}
glimpse(match_1_clean)
```

This dataset, with 157 variables and 3,165 observations, corresponds to the 'Liv_v_Spurs_Event_Data.csv' dataset stored in my [Github repo](https://github.com/buckleco/Portfolio_Data/blob/main/Liv_v_Spurs_Event_Data.csv), which I also use in some other blog posts. 
